Automatically Proving Equivalence by Type-Safe Reflection

### 2017-05-20 paper-CICM-17/content/1-introduction.tex
A bit more related work

The only way for $(\code{Id}_a\ x\ y)$ to be inhabited is therefore that $x$ and $y$ are equal by definition. In this case, the constructor \code{Refl} helps to create a proof of this equality: \code{(Refl\ x)} is precisely the proof
which says that $x=_ax$.

@@

Our handwritten proof \code{adc\_lemma\_2} uses only the existence of a neutral
element, and the associativity and commutativity of $+$ on \code{Nat}. Thus,
we're rewriting a term by using the properties of a commutative monoid. 
With the right choice of combinators, such proofs could be made much simpler,
but we would like a generic prover for commutative monoids to find a proof
automatically.

>>>>

The only way for $(\code{Id}_a\ x\ y)$ to be inhabited is therefore that $x$ and $y$ are equal by definition. In this case, the constructor \code{Refl} helps to create a proof of this equality: \code{(Refl\ x)} is precisely the proof
which says that $x=_ax$. 
Here, we are using the notation of Idris, where unbound variables
like \texttt{a} in the definition of \texttt{Id} are
\emph{implicitly} quantified, as a concise programming notation.

@@

Our handwritten proof \code{adc\_lemma\_2} uses only the existence of a neutral
element, and the associativity and commutativity of $+$ on \code{Nat}. Thus,
we're rewriting a term by using the properties of a commutative monoid. 
With the right choice of combinators~\cite{Carette12} such proofs could be made
much simpler, but we would like a generic prover for commutative monoids to
find a proof automatically.

### 2017-05-20 paper-CICM-17/content/3-hierarchyOfTactics.tex
A bit more related work

A novelty of our work is that instead of building a prover for a
specific algebraic structure (like ring), we have built a hierarchy of provers
for many algebraic structures. Moreover, each prover reuses components
of the others so that
the various simplifications are not duplicated at the different
levels: the normalisation function of each structure uses as much as possible
the normalisation function of the structure from which it inherits. For
example, the normalisation on monoids reuses the normalisation on semi-groups
in such a way that it does not have to deal with associativity. In this
case, the datatype reflecting terms in semi-groups has
the same expressive power as that for monoids, so a term valid in a monoid can easily be transformed into a corresponding term of a semi-group. However, 
there is a difficulty with
groups and monoids: normalisation for groups
needs to reuse the normalisation for monoids, meaning that
we will have to encode 
negations somehow, which can't be directly expressed in a monoid. For this
reason, we develop some specific encodings.

@@

We have built an automatic reflection mechanism which enables the machine to
build the reflected terms for us automatically. This is not essential to our
approach, but it simplifies the usage of the tactics by removing the need to
write long encodings by hand.  To do so, we used Idris' reflection mechanism,
which enables pattern matching on syntax, rather than on constructors. 
While we omit the full details due to space constraints, reflecting values
involves defining functions of the following form:

>>>>

A novelty of our work is that instead of building a prover for a
specific algebraic structure, we have built a hierarchy of provers.
Each prover reuses components
of the others so that
the simplifications are not duplicated at different
levels: normalisation of each structure uses as much as possible
the normalisation function of the structure from which it inherits. For
example, normalisation on monoids reuses normalisation on semi-groups
so that it does not have to deal with associativity. In this
case, the datatype reflecting terms in semi-groups has
the same expressive power as that for monoids, so a term in a monoid can
be transformed directly into a corresponding term of a semi-group. However, 
there is a difficulty with
groups and monoids: if normalisation for groups
uses the normalisation for monoids, 
we will have to encode 
negations somehow, which can't be directly expressed in a monoid.
Therefore, we develop some specific encodings.

@@

We have built an automatic reflection mechanism which enables the machine to
build reflected terms. This is not essential to our
approach, but it simplifies the usage of the tactics by removing the need to
write long encodings by hand.  To do so, we used Idris' reflection mechanism,
which enables pattern matching on syntax, rather than on constructors. 
While we omit the full details due to space constraints, reflecting values
involves defining functions of the following form:

### 2017-05-20 paper-CICM-17/content/4-relatedWork.tex
A bit more related work

Coq's ring solver~\cite{Coq2005}, like ours, is
implemented using proof-by-reflection techniques, but without the guarantees obtained with
our type-safe reflection mechanism, and without the correct-by-construction
approach: first, they define the normalisation of terms, then
they prove correctness of the normalisation with an external lemma:
$\forall\ (e1\ e2\ :\ \code{Expr}),\ beq_{Expr}\ (\code{norm}\ e1)\
(\code{norm}\ e2)\ =\ true\ \rightarrow\ \code{reify}\ e1\ \simeq\
\code{reify}\ e2$.
This needs a \code{reify} function that we do not
in our development. Another difference is that Coq's prover only deals
with rings and semi-rings (commutative or not), but not with any of the
intermediate structures (semi-group, monoid, group, etc), unlike our solver. 
However,
their implementation has better performances due to the use of sparse normal
form and more optimised algorithms.  Our automatic reflection was written with
Idris reflection mechanism which allows pattern matching on syntax, and
their automatic reflection is programmed in Ltac~\cite{DelahayeLTac}, a proof
dedicated and untyped meta-language for the writing of automations.
More recently, the Mtac extension~\cite{Ziliani13} provides a
typed language for implementation proof automations.

Proof automations have been investigated for various proof assistants and
programming languages. In Coq, appart from the ring prover, there's also the
Omega solver~\cite{Cregut04}, which solves a goal in Presburger arithmetic
(i.e. a universally quantified formula made of equations and inequations), and
a field~\cite{DelahayeField} decision procedure for real numbers, which plugs
to Coq's ring prover after simplification of the multiplicative inverses.
Various automations have also been done in
Agda~\cite{DBLP:conf/mpc/KokkeS15,Lindblad04}.
%
Proofs by reflection has been intensively
studied~\cite{ChlipalaBook,Malecha14}, but again without anything similar to
the type-safe reflection that we have presented here.

If we leave the ground of nice mathematical structures, one can decide
to work with arbitrary rewriting rules, but in the general case there isn't a
complete decision procedure for such systems, because there is usually no
normal form. This is where deduction modulo~\cite{Dowek03,DelahayeModulo} and
proof search heuristics start.

>>>>

Coq's ring solver~\cite{Coq2005}, like ours, is
implemented using proof-by-reflection techniques, but without the guarantees obtained with
our type-safe reflection mechanism, and without the correct-by-construction
approach: first, they define the normalisation of terms, then
they prove correctness of the normalisation with an external lemma:
$\forall\ (e1\ e2\ :\ \code{Expr}),\ beq_{Expr}\ (\code{norm}\ e1)\
(\code{norm}\ e2)\ =\ true\ \rightarrow\ \code{reify}\ e1\ \simeq\
\code{reify}\ e2$.
This needs a \code{reify} function, which we do not need.
Furthermore, Coq's prover deals
with rings and semi-rings (commutative or not), but not with any of the
intermediate structures (semi-group, monoid, group, etc).
However,
their implementation has better performances due to the use of sparse normal
form and more optimised algorithms.  Our automatic reflection was written with
Idris reflection mechanism which allows pattern matching on syntax, and
their automatic reflection is programmed in Ltac~\cite{DelahayeLTac}, a proof
dedicated and untyped meta-language for the writing of automations.
More recently, the Mtac extension~\cite{Ziliani13} provides a
typed language for implementating proof automation.

As well as the ring solver, Coq also provides the
Omega solver~\cite{Cregut04}, which solves a goal in Presburger arithmetic
(i.e. a universally quantified formula made of equations and inequations), and
a field~\cite{DelahayeField} decision procedure for real numbers, which plugs
to Coq's ring prover after simplification of the multiplicative inverses.
Agda's reflection
mechanism\footnote{\url{http://wiki.portal.chalmers.se/agda/pmwiki.php?n=ReferenceManual.Reflection}}
gives access to a representation of the current goal (that is,
the required type) at a particular point in a program.
This allows various proof automations to be done in
Agda~\cite{DBLP:conf/mpc/KokkeS15,Lindblad04}.

Proofs by reflection has been intensively
studied~\cite{ChlipalaBook,Malecha14}, but without anything similar to
the type-safe reflection that we have presented here.
%
If we leave the ground of nice mathematical structures, one can decide
to work with arbitrary rewriting rules, but in the general case there isn't a
complete decision procedure for such systems, because there is usually no
normal form. This is where deduction modulo~\cite{Dowek03,DelahayeModulo} and
proof search heuristics start.

### 2017-05-20 paper-CICM-17/content/5-conclusions.tex
A bit more related work

We have implemented a very generic solution to the initial problem of index
mismatch that appeared in section~\ref{sect:motivatingExample} when we were
using indexed types. This solution takes the form of a hierarchy of provers for
equivalences in algebraic structures. These provers are generic in multiple
ways: they work for many algebraic structures (semi-group, monoid, commutative
monoid, group, commutative group, ring and semi-ring), for any type that
behaves as one of these structures, and for any equivalence relation on this
type (and not only for the propositional equality). The implementation is
modular and each prover reuses the prover of the structure from which it
inherits. These provers can automatically prove equivalences between terms,
which discharges the user of providing proofs for some proof obligations, as
\code{adc\_lemma\_2}. Thus, these provers enable the user to focus on the
interesting proofs that requires specific knowledge and creativity, instead of
wasting time for proving routine lemmas that can be automated.

The correct-by-construction method that we have followed has involved the
design of a type-safe reflection mechanism where reflected terms are indexed
over the concrete inputs, and from which we are able to pull out the proofs
easily. The construction of the proof is done step by step, and it follows the
construction of the normalised terms. Therefore, unlike Coq's and Agda's ring
provers, we do not have the traditional duplication that happens when
separating the computational content from the proof of correctness. Instead, in
our approach, the construction of the proof is guided by the construction of
the normalised terms, which, in addition of avoiding redundancy, simplifies the
proof generation considerably. This development shows that if dependent types
effectively bring some new problems, they are also very expressive tools for
building correct-by-construction software where the development is \emph{driven
by the types}.

This work can be extended to build new provers for less common algebraic
structures and for more specific structures. For example, regular expressions
form a ``pre semi-ring" with some extra axioms. We will refactor the semi-ring
level with the creation of the intermediate structure of pre semi-ring, that
will not necessary have the property that $0$ is an annihilator element for the
product. Then, we could build a specific prover for regular expressions, that
would use the normalisation function of the pre semi-ring level and that would
only have to deal with the specific properties of regular expressions : the
neutral element $\varnothing$ for the concatenation of languages is also a
neutral element for the product of languages, the idempotence of the addition
of languages, and the rules of simplifications for the new Kleene star
operation.

>>>>

We have implemented a generic solution to the initial problem of index
mismatch (section~\ref{sect:motivatingExample}) when 
using indexed types. This solution takes the form of a hierarchy of provers for
equivalences in algebraic structures. These provers are generic in several
ways: they work for many algebraic structures (semi-group, monoid, commutative
monoid, group, commutative group, ring and semi-ring); for any type that
behaves as one of these structures; and, for any equivalence relation on this
type. The implementation is
modular and each prover reuses the prover of the structure from which it
inherits. These provers can automatically prove equivalences between terms,
so the user need not prove obligations by hand,
like \code{adc\_lemma\_2}. Thus, these provers enable the user to focus on the
interesting proofs that requires specific knowledge and creativity, instead of
routine, automatable, lemmas.

Our correct-by-construction method involved the design of a type-safe
reflection mechanism where reflected terms are indexed over concrete inputs,
and from which we are able to extract proofs directly.  Unlike Coq's and Agda's
ring provers, we do not have the duplication that arise when separating the
computational content from the proof of correctness. Instead, construction of
the proof is done step by step, following the construction of the normalised
terms.  In addition to avoiding redundancy, this simplifies the proof
generation considerably. This development shows that if dependent types
effectively bring some new problems, they are also very expressive tools for
building correct-by-construction software where the development is \emph{driven
by the types}.

This work can be extended to build new provers for less common algebraic
structures and for more specific structures. For example, regular expressions
form a ``pre semi-ring'' with some extra axioms. We will refactor the semi-ring
level with the creation of the intermediate structure of pre semi-ring, that
will not necessary have the property that $0$ is an annihilator element for the
product. Then, we could build a specific prover for regular expressions, that
would use the normalisation function of the pre semi-ring level and that would
only have to deal with the specific properties of regular expressions : the
neutral element $\varnothing$ for the concatenation of languages is also a
neutral element for the product of languages, the idempotence of the addition
of languages, and the rules of simplifications for the new Kleene star
operation.
